---
title: Responsible Repositories 
authors:
  - name: Nathanael Sheehan
    department:  College of Engineering, Mathematics and Physical Sciences 
    affiliation: Exeter University
    location: University of Exeter, St Germans Road, EX4 7HD Exeter, UK
    email: ns651@exeter.ac.uk
  - name: Sabina Leonelli
    department: EGENIS, Centre For The Study Of Life Sciences
    affiliation: Exeter University
    location: University of Exeter, St Germans Road, EX4 7HD Exeter, UK
    email: s.leonelli@exeter.ac.uk
abstract: |
  In this paper I aruge on the importance of TRUST 
keywords:
  - SARS-CoV2
  - Open Science
  - Data Governance
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
always_allow_html: true
---

# Introduction

The effectiveness of disseminating results promptly, sometimes even before having them formally published – thereby speeding up research - has been extolled by scientific and popular media alike, most evidently in relation to the prompt dissemination of genetic sequencing data from various strains of the SARSCOV-2 virus (an exemplary instance of ‘Open Data’) and the decision by publishing companies to temporarily release all coronavirus-related papers without charges (‘Open Access’). As the United Nations joins the choir of OS supporters with its 2021 recommendation to implement OS worldwide, the OS movement looks well-positioned to
determine the future of post-pandemic research and related policies. This shift in research practice - in conjunction with decreasing costs in data storage - has led to the latest research findings, treatments and protocols on Covid-19 and related topics becoming freely avalalible on the internet. 

Belief in misinformation about COVID-19 poses a potential risk to public health; therefore, scientists play a key role as disseminators of factual and reliable information (Roozenbeek et al., 2020)
In an Open Science context, the TRUST (Transparency, Responsibility, User Focus, Sustainability, and Technology ) principles highlight a set of guidelines to demonstrate the trustworthiness of a digital repository to many of the stakeholders involved, including researchers, community users, funders, developers and service providers. "Trustworthiness is demonstrated through evidence, which depends on transparency, and thus repositories must provide transparent, honest, and verifiable evidence of their practice. In this way, stakeholders can be confident that repositories ensure data integrity, authenticity, accuracy, reliability, and accessibility over extended time frames. Trustworthiness is not a one-off achievement; it cannot be taken for granted without regular audit and certification" (Lin et al 2020)

# Method
propose metric for evaluating "R", as done with FAIR principles

# Discussion

(3) suggest that this metric is insufficient or anyhow ambiguous, because it takes no account of an additional and central factor in assessing responsibility: i.e. underpinning interpretation of what counts as responsible openness

(4) suggest that this should be added as a crucial additional factor for R, BUT also that this cannot be easily transformed into a binary metric (1-0) -- we are looking at qualitative differences

(5) discuss how this works out in case og GISAID vs COVID-19 portal: here we have a clash of ideologies and experiences of what constitutes "good" openness

# Conclusion

(6) conclusion: there is much that CAN be done to metricise and evaluate TRUST principles as a key complement to FAIR, however even such evaluation needs to highlight the unavoidable qualitative/interpretative differences in the implementation of openness

```{r fig1, fig.cap = "R metric table", echo=FALSE, message=FALSE, warning=FALSE}
library(formattable)
r_score = data.frame(matrix(ncol=8, nrow = 2))
cols = c("Repository","R1.1 Complete Metadata","R1.2 Techincal Documentation",
         "R1.3 Quality Control", "R1.4 Authenticity Protection",
         "R2.1 Reliable Data Services", "R3.1 Manage IP of Data Producers",
         "R3.2 Security of System & Content"
)
colnames(r_score) = cols


# Complete score for GISAID
r_score$Repository[1] = "GISAID"
r_score$`R1.1 Complete Metadata`[1] = 0.5
r_score$`R1.2 Techincal Documentation`[1] = 0.5
r_score$`R1.3 Quality Control`[1] = 1.0
r_score$`R1.4 Authenticity Protection`[1] = 1.0
r_score$`R2.1 Reliable Data Services`[1] = 1.0
r_score$`R3.1 Manage IP of Data Producers`[1] = 1.0
r_score$`R3.2 Security of System & Content`[1] = 0.5


# Complete score for Covid-19 Data Platform
r_score$Repository[2] = "Covid-19 Data Platform"
r_score$`R1.1 Complete Metadata`[2] = 0.5
r_score$`R1.2 Techincal Documentation`[2] = 0.5
r_score$`R1.3 Quality Control`[2] = 1.0
r_score$`R1.4 Authenticity Protection`[2] = 0
r_score$`R2.1 Reliable Data Services`[2] = 0.5
r_score$`R3.1 Manage IP of Data Producers`[2] = 0
r_score$`R3.2 Security of System & Content`[2] = 0.5

# Sum results
r_score$Score[1] = sum(r_score[1,2:8])
r_score$Score[2] = sum(r_score[2,2:8])

customGreen0 = "#DeF7E9"
customGreen = "#71CA97"
customRed = "#ff7f7f"
library(DT)

formattable(r_score, align =c("l","c","c","c","c", "c", "c", "c", "r"), list(
  `Repository` = formatter("span", style = ~ style(color = "grey",font.weight = "bold")),
  `R1.1 Complete Metadata`= color_tile(customGreen0, customGreen0),
  `R1.2 Techincal Documentation`= color_tile(customGreen0, customGreen0),
  `R1.3 Quality Control`= color_tile(customGreen, customGreen),
  `R1.4 Authenticity Protection`= color_tile(customRed, customGreen),
  `R2.1 Reliable Data Services`= color_tile(customGreen0, customGreen),
  `R3.1 Manage IP of Data Producers`= color_tile(customRed, customGreen),
  `R3.2 Security of System & Content`= color_tile(customGreen, customGreen0),
  `Score`= color_tile(customRed, customGreen)
))

```


Albiet, TRUST is not a new concept and a number of trustworthiness certification mechanisms already exist; The Open Archival Information System (OAIS) provides a recommendation model to provide long-term preservation and access to digital information[REF]. The FAIR principles emphasize a best practice of machine and human re-usability with data objects.  

<!-- when defining key qualities of data objects that are  -->

<!--  believe thr prinicples are the -->

<!-- digital repository trustworthiness.  -->

<!-- The biological entity scietifically known  -->
<!-- Among the vast digital sea of data reposotories the usual authorities who steer the direction of power and resources are present, Governments and Universities were among the first to adopt OS principles in practice, daily updates of cases testing, vaccine progress, epistomeoglocial roots interali    -->

<!-- it takes a great deal of work to sucessfully collect, store and manage data  -->
<!-- and among the hundreds of thousands of reposotoires which have come online since the pandemic -->

<!-- in cyberspace and are owned and mangaged by an array of Governments, Universities, Businesses and publics.  -->


<!-- Data repostiores have been buuilt by Governments, univieristies, busisinerses and publics -->
<!-- The greatest example of this has been the sharing of SARS genomic data, where millions of genomes have been shared across the world  -->
<!-- During the Pandemic an open letter was written by the EBI in support in complete openness in genome sharing and asked for people to donate all data to a traid of databases -->
<!-- Many public figures and institutions supported the letter saying that coivd was bigger than all of us, and we must come together -->
<!-- Others critiseied the letter and argued that there database was a wolf in open source clothing and missed the point in resonsibility for data sharing -->




<!-- TRUSt is a set of principles for data repositiores owners to adhere to so the community is served and they demonstate the ability to mananage the data they hold -->
<!-- T = transparancey U = User focus, S = Sustainablity, T = Trust -->
<!-- R = Adhering to the designated community’s metadata and curation standards, along with providing stewardship of the data holdings e.g. technical validation, documentation, quality control, authenticity protection, and long-term persistence. -->

<!-- Providing data services e.g. portal and machine interfaces, data download or server-side processing. -->

<!-- Managing the intellectual property rights of data producers, the protection of sensitive information resources, and the security of the system and its content. -->

<!-- The FAIR Data Principles3 highlight the need to embrace good practice by defining essential characteristics of data objects to ensure that data are reusable by humans and machines  -->
<!-- The Open Archival Information System (OAIS) reference model4 provides recommendations on setting up archives delivering long-term preservation of and access to information (in particular, digital information) and creating preservation packages. -->
<!-- None of these encapsulate trust and tremporal demand that trust takes to build and serve a community -->
<!-- [trust table ] -->

<!-- To start I document my expereince joining each community and qualiotaitely narrate me expereince conceptaulising the R principles in each repo -->
<!-- I then systhensis the datasets into aggreated months and explore statistical trends in each dataset -->
<!-- I conclusde by a final discourse on how  -->

# Methods

```{r fig2, fig.cap = "Monthly totals of global SARS-CoV-2 cases sequenced and shared\n on the GISAID and Covid-19 Data Platform database until Febuary 22 2022", echo=FALSE, message=FALSE, warning=FALSE}
source("../code/set-up.r")
main_df = readRDS("../data/main_df.rds")
temporal_sub_all = readRDS("../data/temporal_sub_all.rds")
ggplot(data=temporal_sub_all) +
  geom_line(aes(x=Date,y=sum_gisaid, group=1, color='GISAID Monthly Total')) +
  geom_line(aes(x=Date,y=sum_cd19dp, group=1, color ='CD19DP Monthly Total')) +
  scale_colour_manual("",
                      breaks = c("GISAID Monthly Total", "CD19DP Monthly Total"),
                      values = c("green", "blue")) +
  labs(x = "Date",
       y = "Sequence Submissions",
       title = "Monthly Total SARS-CoV-2 Sequence Submissions",
       caption = "GISAID Metadata: https://www.epicov.org/\nCovid-19 Data Platform Metadata: https://www.ebi.ac.uk/ena/portal/api/ ") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5),   legend.position = "bottom")


```






