---
title: Responsible Repositories 
authors:
  - name: Nathanael Sheehan
    department:  College of Engineering, Mathematics and Physical Sciences 
    affiliation: Exeter University
    location: University of Exeter, St Germans Road, EX4 7HD Exeter, UK
    email: ns651@exeter.ac.uk
  - name: Sabina Leonelli
    department: EGENIS, Centre For The Study Of Life Sciences
    affiliation: Exeter University
    location: University of Exeter, St Germans Road, EX4 7HD Exeter, UK
    email: s.leonelli@exeter.ac.uk
abstract: |

keywords:
  - SARS-CoV2
  - Open Science
  - Data Governance
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
always_allow_html: true

---

# Introduction

The pandemic demanded rapid, scalable and open access to the latest research findings, treatments and protocols on the coronavirus. This shift in research practice - in conjunction with decreasing costs in data storage - has led to an exponential increase in data repositories becoming publicly accessible online, as well as a global recognition of Open Science (OS) principles by the United Nations [REF]. However, during the same time, a wider discourse emerged among policy circles and researchers with concern to the responsibility of repositories reporting on COVID-19 and related topics, most evidently in relation to the prompt dissemination of genetic sequencing data from various strains of the SARS COV-2 virus. 

On the 29th of January 2021 the governing board of the European Bioinformatics Institute (EBI) posted a public letter in Nature calling for a greater "openness" in sharing SARS-CoV-2 genome data, the letter argued "to unleash the fast flow of research advances" the scientific community must remove all formal barriers which restrict data sharing and share all genome data to one of a triad of state databases (EBI, The Gene Bank USA, Japan Gene repo). At the same time, the Global Initiative on Sharing Avian Influenza Data (GISAID) EpiCov database had just overtaken the EBI's European COVID-19 Data Portal (C19DP) in the volume of genome sequences being shared. GISAID was originally established in 2008 to monitor global influenza outbreaks and is now hosted by the German Institute of Agriculture and the Max Planck institute; the platforms pivot quickly grew in popularity during the pandemic and received a number of philanthropic donations from the World Health Organization, big-pharma and nation states, unlike C19DP, GISAID requires that users confirm their identity and agree not to republish the sites genomes without permission from the data provider.

@pittphilsci19817 introduces this case study as a clash in "responsible research measures", pointing out how "trustworthy and explicitly non-exploitative conditions for data sharing helps to widen participation in data sharing", in this same line of thought we stipulate that Trust and Responsibility, or a lack of it, plays a central role in the success of OS platforms; hence researchers and data producers alike, must ensure quality and integrity of data when practicing OS principles in order to demonstrate the quality and discipline which OS research demands. 

The TRUST (Transparency, Responsibility, User Focus, Sustainability, and Technology) principles put forth a set of guidelines to  demonstrate the trustworthiness of a digital repository to many of the stakeholders involved, including researchers, community users, funders, developers and service providers. In an OS context, a number of trustworthiness certification mechanisms already exist; The Open Archival Information System (OAIS) provides a recommendation model to provide long-term preservation and access to digital information [REF] and the FAIR principles emphasize a best practice of machine and human re-usability with data objects. However, Lin et al (2020) argue these frameworks are not suitable for a broader audience and lack a critical understanding of the temporal aspects of a data repository, data may enter a system under a FAIR or OAIS system, however TRUST should not be understood as a single check box, rather it must regularly audited to ensure the system ensures trust for the long run (Lin et al 2020). Repository users should have confidence that data depositors are prompted to provide all metadata compliant with the community norms, as this greatly enhances the discoverability and usefulness of the data. Knowing that a repository verifies the integrity of available data and metadata assures potential users that the data holdings are more likely to be interoperable with other relevant datasets. Both depositors and users must have confidence that the data will remain accessible over time, and thus can be cited and referenced in scholarly publications.  Building on previous discussions on Epistemic diversity and Open Science, we propose a novel metric system based on the description of the R principle in TRUST (Lin et al 2020) to be used in tandem with a qualitative understanding to interpret the philosophical nuances of differing implementations of openness.

# Methods

## Definition
Stakeholders of a repository must take responsibility - legally and ethically- for the data being stored and collected from their user community. Lin et al (2020) delineates responsibility being demonstrated as:

- "*Adhering to the designated community’s metadata and curation standards, along with providing stewardship of the data holdings e.g. technical validation, documentation, quality control, authenticity protection, and long-term persistence.*

- *Providing data services e.g. portal and machine interfaces, data download or server-side processing.*

- *Managing the intellectual property rights of data producers, the protection of sensitive information resources, and the security of the system and its content.*"

We classify our reading of responsibility into three distinct modes which are described by sub classifications metrics. Each metric varies on a scale from 0-1, where 0 indicates the responsibility classification is "not implemented", 0.4 indicates the responsibility classification is "partially implemented" and 0.9 indicates the responsibility classification is "Sufficiently implemented". The total score for each distinct mode is calculated as a mean weighted average from each of its sub classifications. 
 

```{r fig1, fig.cap = "Table of Responsiblity Metric", echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
r_score = data.frame(matrix(ncol = 3, nrow = 5))

cols = c("Metadata Responsibility",
         "Service Responsibility",
         "Legal Responsibility")
colnames(r_score) = cols

mr = c(  "R1.1 Curated Metadata",
         # metadata which documents the progress of data in the repo to date
         "R1.2 Techincal Documentation",
         # documentation online on how to use data services
         "R1.3 Format Control",
         # uploading data
         "R1.4 Content Control",
         # peer review
         "R1.5 Authenticated Provinence")
r_score$`Metadata Responsibility` = mr

sr = c("R2.1 Reliable Data Services")
r_score$`Service Responsibility`[1] = sr

lr = c(  # LEGAL Responsibility
  "R3.1 Manage IP of Data Producers",
  # legal agreement much can be done
  "R3.2 Security of System",
  # does the system have vulnerabilities
  "R3.3 Security of Content" # how is content ensured its not reshared ?
  )
r_score$`Legal Responsibility`[1:3] = lr
r_score[is.na(r_score)] = ""   
kable(r_score, caption = "Modes of responsiblity and sub classifcations metrics") %>% kable_material_dark() %>%
  kable_styling(latex_options = "HOLD_position")

```

## Metadata Responsibility
Metadata Responsibility is the largest group of sub-classifications and focuses on the quality, re-usability and interoperability of data stored on the repository. `R1.1 Sufficient Metadata` asserts repository metadata to conform to a general curation standard whereby data can be transformed for further discoverability. `R1.2 Technical Documentation` requires a closed reading of each repository 'docs', and evaluates whether the documents provide clear and useful instructions to successfully interact and use the platform. `R1.3 Format Control` concerns a format control test to ensure that users uploading data have to do so in a standardized form. `R1.4 Content Control` asks if user content goes under a form of peer review to ensure integrity of data producers. `R1.5 Authenticated Provinence` checks if there are forms of authentication to interact with the data respiratory. 


## Service Responsibility
Service Responsibility is the smallest group of sub-classifications and focuses on the tools presented to download and upload data. `R2.1 Reliable Data Services` tests the range of tools available to download data from the platform and tests if data can be downloaded in a range of formats and is machine readable. 


## Legal Responsibility
Legal Responsibility is a critical component to the R in trust as it binds the digital and physical world. `R3.1 Manage IP of Data Producers` verifies the intellectual property policy for data producers and the effects this has for  users of the platform. `R3.2 Security of System` studies the vulnerabilities in infrastructure which uphold the data, the platform and the various microservices which contribute to the functioning of the whole system. `R3.3 Security of Content` 

# Results

```{r fig3, fig.cap = "Table of Results", echo=FALSE, message=FALSE, warning=FALSE}

r_score1 = data.frame(matrix(ncol = 3, nrow = 10))

r_score_rn = c("Metric", "GISAID: EpiCov", "EBI: Covid-19 Data Platform")
colnames(r_score1) = r_score_rn

metric_col = c(
  # DATA Responsibility
  "R1.1 Sufficent Metadata",
  # metadata which documents the progress of data in the repo to date
  "R1.2 Techincal Documentation",
  # documentation online on how to use data services
  "R1.3 Format Control",
  # uploading data
  "R1.4 Content Control",
  # peer review
  "R1.5 Authenticated Provinence",
  # log in
  # SERVICES Responsibility
  "R2.1 Reliable Data Services",
  # a range of data tools which permit the download of data in various formats
  # LEGAL Responsibility
  "R3.1 Manage IP of Data Producers",
  # legal agreement much can be done
  "R3.2 Security of System",
  # does the system have vulnerabilities
  "R3.3 Security of Content" ,
  # how is content ensured its not reshared ?
  "Total Score"
  
)
r_score1$Metric = metric_col

r_score1$`EBI: Covid-19 Data Platform`[1] = 0.9
r_score1$`EBI: Covid-19 Data Platform`[2] = 0.9
r_score1$`EBI: Covid-19 Data Platform`[3] = 0.9
r_score1$`EBI: Covid-19 Data Platform`[4] = 0.9
r_score1$`EBI: Covid-19 Data Platform`[5] = 0
r_score1$`EBI: Covid-19 Data Platform`[6] = 0.4
r_score1$`EBI: Covid-19 Data Platform`[7] = 0
r_score1$`EBI: Covid-19 Data Platform`[8] = 0.4
r_score1$`EBI: Covid-19 Data Platform`[9] = 0
r_score1$`EBI: Covid-19 Data Platform`[10] = sum(r_score1$`EBI: Covid-19 Data Platform`[1:9])

r_score1$`GISAID: EpiCov`[1] = 0.4
r_score1$`GISAID: EpiCov`[2] = 0.4
r_score1$`GISAID: EpiCov`[3] = 0.9
r_score1$`GISAID: EpiCov`[4] = 0.9
r_score1$`GISAID: EpiCov`[5] = 0.9
r_score1$`GISAID: EpiCov`[6] = 0.9
r_score1$`GISAID: EpiCov`[7] = 0.4
r_score1$`GISAID: EpiCov`[8] = 0.4
r_score1$`GISAID: EpiCov`[9] = 0.9
r_score1$`GISAID: EpiCov`[10] = sum(r_score1$`GISAID: EpiCov`[1:9])


kable(r_score1, caption = "Results", format = "latex")  %>% kable_material_dark() %>%
  kable_styling(latex_options = "HOLD_position")



```

# Discussion

<!-- This paper has presented a novel but highly subject approach  -->
<!-- The RR metric is a novel approach to quantitatively study the responsibility of a data repository, however is   -->

<!--  to truly reflect a robust description of repository responsibility. A multitude of factors were not considered in this analysis, such as  -->
<!-- how practitioners of openness interpret what counts as responsible openness or how community users interact and use each platform. Further work refining the binary metric system may serve as a robust piece of empirical contribution in understanding the general question of responsibility in respotistiores, however only when coupled with qualitative differences.  -->


<!-- (4) suggest that this should be added as a crucial additional factor for R, BUT also that this cannot be easily transformed into a binary metric (1-0) -- we are looking at  -->

<!-- (5) discuss how this works out in case og GISAID vs COVID-19 portal: here we have a clash of ideologies and experiences of what constitutes "good" openness -->

<!-- # Conclusion -->

<!-- (6) conclusion: there is much that CAN be done to metricise and evaluate TRUST principles as a key complement to FAIR, however even such evaluation needs to highlight the unavoidable qualitative/interpretative differences in the implementation of openness -->

# References

```{r fig2, fig.cap = "Monthly totals of global SARS-CoV-2 cases sequenced and shared\n on the GISAID and Covid-19 Data Platform database until Febuary 22 2022", echo=FALSE, message=FALSE, warning=FALSE}
source("../code/set-up.r")
main_df = readRDS("../data/main_df.rds")
temporal_sub_all = readRDS("../data/temporal_sub_all.rds")
ggplot(data=temporal_sub_all) +
  geom_line(aes(x=Date,y=sum_gisaid, group=1, color='GISAID Monthly Total')) +
  geom_line(aes(x=Date,y=sum_cd19dp, group=1, color ='CD19DP Monthly Total')) +
  scale_colour_manual("",
                      breaks = c("GISAID Monthly Total", "CD19DP Monthly Total"),
                      values = c("green", "blue")) +
  labs(x = "Date",
       y = "Sequence Submissions",
       title = "Monthly Total SARS-CoV-2 Sequence Submissions",
       caption = "GISAID Metadata: https://www.epicov.org/\nCovid-19 Data Platform Metadata: https://www.ebi.ac.uk/ena/portal/api/ ") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5),   legend.position = "bottom")


```










