---
title: "OSS dashboards: Epistemic Diversity"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    social: menu
    source: embed
---
<style>                     
.navbar {
  background-color:#cf2e2e;
  border-color:black;
}
.navbar-brand {
color:black!important;
}
</style> 

```{r setup, include=FALSE}
source("set-up.r")
main_df = readRDS("../data/main_df.rds")
temporal_sub_all = readRDS("../data/temporal_sub_all.rds")
cd19_agg_data = readRDS("../data/cd19_agg_data.RDS")
gisaid = readRDS("../data/gisaid.RDS")
k3 = readRDS("../data/k3_plot.RDS")
k9 = readRDS("../data/k9_plot.RDS")
```

### Fig. 1: Monthly totals of global SARS-CoV-2 cases sequenced and shared on the GISAID and Covid-19 Data Platform database until Febuary 22 2022.

```{r}
f1= ggplot(data=temporal_sub_all) +
  geom_line(aes(x=Date,y=sum_gisaid, group=1, color='GISAID Monthly Total')) +
  geom_line(aes(x=Date,y=sum_cd19dp, group=1, color ='CD19DP Monthly Total')) +
  scale_colour_manual("",
                      breaks = c("GISAID Monthly Total", "CD19DP Monthly Total"),
                      values = c("green", "blue")) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1)) +
  labs(x = "Date",
       y = "Sequence Submissions",
       title = "Monthly Total SARS-CoV-2 Sequence Submissions",
       caption = "") + dark_theme() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
ggplotly(f1)

```

***

GISAID Metadata: 

- https://www.epicov.org/

Covid-19 Data Platform Metadata:

- https://www.ebi.ac.uk/ena/portal/api/ 



### Fig. 2:  Monthly mean of global SARS-CoV-2 cases sequenced and shared on the GISAID and Covid-19 Data Platform database until Febuary 22 2022.

```{r}
f2 = ggplot(data=temporal_sub_all) +
geom_line(aes(x=Date,y=sum_gisaid_av, group=1, color='GISAID Monthly Mean')) +
  geom_line(aes(x=Date,y=sum_cd19dp_av, group=1, color ='CD19DP Monthly Mean')) +
  annotate("text", x = 7, y = 600, label = "WHO Declares\n Pandemic") +
  annotate("text", x = 17, y = 2000, label = "AstraZeneca's \nVaccine Authorised") +
  annotate("text", x = 18, y = 4000, label = "EBI Open letter") +
  annotate("text", x = 27, y = 6000, label = "Global Covid-19 Deaths \nPass Five Million") +
  scale_colour_manual("",
                      breaks = c("GISAID Monthly Mean", "CD19DP Monthly Mean"),
                      values = c("green", "blue")) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1)) +
  labs(x = "Date",
       y = "Sequence Submissions",
       title = "Monthly Mean SARS-CoV-2 Sequence Submissions")  + dark_theme() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
ggplotly(f2)

```

***
GISAID Metadata: 
 
 - https://www.epicov.org/
 
 Covid-19 Data Platform Metadata: 
 
 - https://www.ebi.ac.uk/ena/portal/api/ 



### Fig. 3: The proportion of global SARS-CoV-2 cases sequenced and shared on the GISAID database until Febuary 22 2022.

```{r}
log_df = main_df %>% filter(Date == "2022/02")

f3 = ggplot(log_df, aes((GISAID.total.Submissions), (log_df$cases))) +
  geom_point(aes(color = continent, size = `Genomes per confirmed cases (GISAID)`),alpha = 75 /100) +
  geom_text(aes(label = Country, color = continent),
            nudge_y = 0.06) + dark_theme() +
  labs(x = "Sequenced Genomes",
       y = "Confirmed Cases",
       title = "Confirmed Covid-19 Cases vs. Sequenced Genomes in the GISAID Open database",
       caption = "\n\n\nSARS-CoV-2 Data: John Hopkins")  +
  scale_y_log10() +
  scale_x_log10() +
  geom_rug(col =rgb(.5, 0, 0, alpha = .2)) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1))

ggplotly(f3)
```

***


### Fig. 4: The proportion of people fully vaccinated and global SARS-CoV-2 cases sequenced and shared on the GISAID database until Febuary 28 2022.

```{r}
f4 = ggplot(log_df, aes((GISAID.total.Submissions),(log_df$People_fully_vaccinated))) +
  geom_point(aes(color = continent, size = `Genomes per confirmed full vaccine (GISAID)`),alpha = 75 /100) +
  geom_text(aes(label = Country,color = continent),nudge_y = 0.06) +
  dark_theme() +
  labs(x = "Sequenced Genomes",
       y = "People Fully Vaccinated",
       title = "People Fully Vaccinated vs. Sequenced Genomes in the GISAID Open Database")  +
  scale_y_log10() +
  scale_x_log10() +
  geom_rug(col =rgb(.5, 0, 0, alpha = .2)) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1))

ggplotly(f4)
```

***



### Fig. 5: The proportion of global SARS-CoV-2 cases sequenced and shared on the C19DP database until Febuary 28 2022

```{r}
f5 = ggplot(log_df, aes(CD19DP.total.Submissions, cases)) +
  geom_point(aes(color = continent, size = `Genomes per confirmed cases (C19DP)`),
             alpha = 75 / 100) +
  geom_text(aes(label = Country, color = continent), nudge_y = 0.06) +
  dark_theme() +
  labs(
    x = "Sequenced Genomes",
    y = "Confirmed Cases",
    title = "Confirmed Covid-19 Cases vs. Sequenced Genomes in the Covid-19 Data Portal"
  )  +
  scale_y_log10() +
  scale_x_log10() +
  geom_rug(col = rgb(.5, 0, 0, alpha = .2)) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1))

ggplotly(f5)
```

***



### Fig. 6: The proportion of people fully vaccinated and global SARS-CoV-2 cases sequenced and shared on the C19DP database until Febuary 28 2022

```{r}
f6 = ggplot(log_df, aes((CD19DP.total.Submissions),
                              (log_df$People_fully_vaccinated))) +
  geom_point(aes(size = `Genomes per confirmed full vaccine (C19DP)`, color = continent),
             alpha = 75 / 100) + geom_text(aes(label = Country, color = continent), nudge_y = 0.06) +
  dark_theme() +
  labs(
    x = "Sequenced Genomes",
    y = "People Fully Vaccinated",
    title = "People Fully Vaccinated vs. Sequenced Genomes in the Covid-19 Data Portal"
  )  +
  scale_y_log10() +
  scale_x_log10() +
  geom_rug(col = rgb(.5, 0, 0, alpha = .2)) +
  guides(shape = guide_legend(order = 2), col = guide_legend(order = 1))

ggplotly(f6)
```

***


### Fig. 7 Covid-19 Data Platform Temporal

```{r}
cd19_agg_data %>%
  barChartRace(
    x = "CD19DP.total.Submissions",
    y = "Country",
    time = "Date",
    ytitle = "Country",
    xtitle = "Count (n submissions)",
    title = "Global GISAID EpiCov Database Submissions",
    paddingWidth = 0.1,
    xFontSize = 10,
    yFontSize = 10,
    xticks = 12,
    xtitleFontSize = 14,
    ytitleFontSize = 14,
    titleFontSize = 22,
    stroke = "black",
    strokeWidth = NULL,
    font = "gochi",
    bgcol = "#cf2e2e",
    panelcol = "#fcb900",
    xgridlinecol = "#8ed1fc",
    opacity = 1,
    timeLabel = TRUE,
    timeLabelOpts = list(
      size = 28,
      prefix = "",
      suffix = "",
      xOffset = 0.5,
      yOffset = 1
    ),
    width = NULL,
    height = NULL
  )
```

***


### Fig. 8 GISAID Temporal Submissions.

```{r}
gisaid %>%
  barChartRace(
    x = "GISAID.total.Submissions",
    y = "Country",
    time = "Date",
    ytitle = "Country",
    xtitle = "Count (n submissions)",
    title = "Global GISAID EpiCov Database Submissions",
    paddingWidth = 0.1,
    xFontSize = 10,
    yFontSize = 10,
    xticks = 12,
    xtitleFontSize = 14,
    ytitleFontSize = 14,
    titleFontSize = 22,
    stroke = "black",
    strokeWidth = NULL,
    font = "gochi",
    bgcol = "#cf2e2e",
    panelcol = "#fcb900",
    xgridlinecol = "#8ed1fc",
    opacity = 1,
    timeLabel = TRUE,
    timeLabelOpts = list(
      size = 28,
      prefix = "",
      suffix = "",
      xOffset = 0.5,
      yOffset = 1
    ),
    width = NULL,
    height = NULL
  )
```

***



### Fig. 9 Clustering (4)

```{r}
ggplotly(k3)

```

***

### Fig. 9 Clustering (9)

```{r}
ggplotly(k9)

```

***

***
### Fig.10 GISAID Share
```{r}
library(treemap)
present = main_df %>% dplyr::filter(Date == "2022/02")

treemap(present,
        index=c("continent", "Country"),
        vSize="GISAID.total.Submissions",
        type="index",
        title = "GISAID EpiCov global \nsequence share",
        fontfamily.title = "mono",
        fontsize.title = 24,
        fontsize.labels=c(12,9),
        fontcolor.labels=c("white","black"),
        fontface.labels=c(2,1),
        bg.labels=c("transparent"),
        palette="RdYlGn",
        align.labels=list(
          c("center", "center"),
          c("right", "top")),
        overlap.labels=0.5,
        inflate.labels=F,
)
```

***
### Fig.10 EBI Global Share
```{r}
library(treemap)
present = main_df %>% dplyr::filter(Date == "2022/02")

treemap(present,
        index=c("continent", "Country"),
        vSize="CD19DP.total.Submissions",
        type="index",
        title = "EBI EpiCov \nglobal sequence share",
        fontfamily.title = "mono",
        fontsize.title = 24,
        fontsize.labels=c(12,9),
        fontcolor.labels=c("white","black"),
        fontface.labels=c(2,1),
        bg.labels=c("transparent"),
        palette="RdYlGn",
        align.labels=list(
          c("center", "center"),
          c("right", "top")),
        overlap.labels=0.5,
        inflate.labels=F,
)
```

***
